---
title: "Functions & Glossary"
author: "University of Nebraska at Kearney Biology"
format: html
editor: visual
---

## Rounding & Formatting

<details>

<summary>Click for information on rounding</summary>

As a reminder, we are following the rounding rules of the *Auk* and *Condor* (now *Ornithology* and *Ornithological Applications*).

Specifically, these rules are as follows:

-   **No naked decimals** (except for caliber)**:**

    -   Preface all decimals with a leading zero; $0.002$ is correct, $.002$ is incorrect.

    -   For caliber, such as with weapons, never have a leading zero; ex., .45 bullet.

-   **Use the same number of decimals as the original data or the precision of the instrument:**

    -   Ex., if precision is $0.01$, then all values should be rounded to hundredths.

-   **Use the same number of decimals for values and uncertainties:**

    -   Ex., $45 \pm 4$, $45.2 \pm 3.8$.

-   **Round percentages to whole numbers\* (or to one decimal):**

    -   For this class, can round to one decimal, but in the referenced publications *always* round to whole number. You will never miss points for rounding percentage to a whole number.

-   **All test statistics should be rounded to two (2) decimal places:**

    -   $F$, $t$, $p$, $\tau$, $\chi^2$, etc. should all be rounded to two decimal places.

-   **All test statistics should show *degrees of freedom*:**

    -   For values with *one* degree of freedom, list like $t_{(9)}=1.56$.

    -   For values with multiple, list them; for ANOVA, put the degrees of freedom for Sum of Squares Between before degrees of freedom for Degrees of Error Within, like $F_{(5,43)}=5.23$.

-   **Round *P*-values as follows:**

    -   $P \ge 0.01$: round to ***two*** decimal places like $P = 0.24$.

    -   $0.01 > P \ge 0.001$: round to ***three*** decimal places like $P = 0.005$.

    -   $0.001 > P \ge 0.0001$: round to ***four*** decimal places like $P = 0.0007$.

    -   $P < 0.0001$: list *exactly* as $P < 0.0001$.

        -   **DO NOT** list without spaces, or as $P<.0001$

</details>

## Common Commands

```{r, echo = F, eval = T, message = F}
library(tidyverse)
library(agricolae)
```

The following are common useful commands used in *R*, with examples of their use.

Commands are given as Windows & Linux / Mac.

### The basics

-   Save your file with `CTRL` + `S` / ⌘ + `S`

-   Undo your last action with `CTRL` + `Z` / ⌘ + `Z`

-   Cut text with `CTRL` + `X` / ⌘ + `X`

-   Copy text with `CTRL` + `C` / ⌘ + `C`

-   Paste text with `CTRL` + `V` / ⌘ + `V`

-   Select all text in a document with `CTRL` + `A` / ⌘ + `A`

### Code basics

-   A code chunk can be inserted using `CTRL` + `SHIFT` + `I` / ⌘ + `shift` + `I`

-   Run a line of code with `CTRL` + `ENTER` / ⌘ + `return`

-   Run a code chunk with `CTRL` + `SHIFT` + `ENTER` / ⌘ + `shift` + `return`

-   Run all code in the document above where you were `CTRL` + `SHIFT` + `ALT` +`P` / ⌘ + `shift` + ⌥ + `P`

-   Run all code with `CTRL` + `ALT` + `R` / ⌘ + ⌥ + `R`

### Code formatting

-   `<-` - save a value as an object. On Mac, the keyboard shortcut is ⌥ + `-`. Windows can be formatted so that the shortcut is `ALT` + `-`.

```{r}
x <- 10
x
```

-   `|>` - "pipe" a command or output into another command. You can use the shortcut `CTRL`+`SHIFT`+`M` / \^ + `SHIFT` + `M`.

```{r}
# make repeatable
set.seed(930)

# random string
x <- rnorm(20)

x |> 
  # pass to summary
  summary() |> 
  # pass summary through round
  round(2)
```

-   `->` - save a value as an object at the end of a pipeline

```{r}
# same pipeline as previous, but more sequential

# get random values
rnorm(20) |> 
  # pass to summary
  summary() |> 
  # pass summary through round
  round(2) ->
  # save as x
  x

x
```

-   `c` - concatenate, place two values together

```{r}
x <- c(10,11)
x
```

### Boolean logic

In order to create arguments in *R*, such as commands like `which`, we need to use Boolean logic commands.

For this example, I will create a string from 1 to 10 to perform the commands on. *Note* that I am using brackets `[ ]` to call out an index within the object using these commands.

These are as follows:

```{r}
# example string
x <- 1:10
```

-   `>`: greater than

```{r}
x[x > 5]
```

-   `<`: less that

```{r}
x[x < 5]
```

-   `>=`: greater than or equal to

```{r}
x[x >= 5]
```

-   `<=`: less than or equal to

```{r}
x[x <= 5]
```

-   `==`: is equal to

```{r}
x[x == 5]
```

-   `|`: or

```{r}
x[x < 3 | x > 7]
```

-   `&`: and

```{r}
x[x > 3 & x < 7]
```

## Basic statistics

<details>

<summary>Click here for basic statistics</summary>

For these examples, we will create a random vector of number to demonstrate how they work.

```{r}
# make repeatable
set.seed(8675309)
# get random normal values
x <- rnorm(1000)
```

-   `mean` - get the mean / average of a set of data

```{r}
mean(x)
```

```{r}
median(x)
```

```{r}
IQR(x)
```

```{r}
range(x)
```

```{r}
# calculate range
abs(range(x)[1] - range(x)[2])
```

```{r}
# calculate range
max(x) - min(x)
```

```{r}
hist(x)
```

</details>

## Null hypotheses

<details>

<summary>Click here for null hypotheses</summary>

### $\chi^2$ tests

#### $\chi^2$ goodness-of-fit

$H_0$: The frequency of observations fits the expected observation frequencies.

#### $\chi^2$ test of independence (i.e., association)

$H_0$: Variable X and Variable Y are not related.

### One-sample tests

#### One-sample $t$-test

$H_0$: $\mu=X$

#### Paired $t$-test

$H_0$: $\mu_1-\mu_2=0$ or $\mu_d=0$

#### Two-sample $t$-test

$H_0$: $\mu_1=\mu_2$

### Multi-sample tests

All ANOVA tests have essentially the same null hypotheses, with the primary null hypothesis being that all means are the same. Secondary hypotheses may exist about *which* means are different; we have to do a post-hoc test to determine which means are different.

$H_0$: $\mu_1=\mu_2=\mu_3=...=\mu_n$

### Correlation

For all correlation tests, the null is that the correlation is 0 (i.e., not correlated positively or negatively). You can get the $p$ value for a test from a $t$ distribution.

As an example, Pearson's correlation:

$H_0$: $\rho = 0$

### Regression

$H_0$: $\beta = 0$

where $\beta$ is the slope.

</details>

## Custom functions from class and elsewhere

The following functions are those developed for this class or adapted from code posted to sources like StackOverflow.

For these stats, we are using the following example data:

```{r}
### EXAMPLE DATA ###
x <- c(1,2,3,5,5,4,6,8,7,9,5)
```

#### Mode calculations

<details>

<summary>Click here for mode</summary>

```{r}
# Based on Statology function
# define function to calculate mode
# works on vectors of data
find_mode <- function(x) {
  # get unique values from vector
  u <- unique(x)
  # count number of occurrences for each value
  tab <- tabulate(match(x, u))
  
  # if no mode, say so
  if(length(x)==length(u[tab == max(tab)])){
    print("No mode.")
  }else{
    # return the value with the highest count
    u[tab == max(tab)]
  }
}

find_mode(x)
```

#### Standard error

```{r}
se <- function(x){
  n <- length(x) # calculate n
  s <- sd(x) # calculate standard deviation
  se_val <- s/sqrt(n)
  return(se_val)
}

se(x)
```

</details>

#### Coefficient of variation

<details>

<summary>Click here for coefficient of variation</summary>

```{r}
cv <- function(x){
  sigma <- sd(x)
  mu <- mean(x)
  val <- sigma/mu*100
  return(val)
}

cv(x)
```

</details>

### Normal distributions

#### $Z$ score

<details>

<summary>Click here for $Z$-score</summary>

Remember - in the $Z$-score code below, if no $n$ is specified, then it will default to $n = 1$.

```{r}
zscore <- function(xbar, mu, sd.x, n = 1){
  z <- (xbar - mu)/(sd.x/sqrt(n))
  return(z)
}

zscore(xbar = 62,
       mu = 65,
       sd.x = 3.5,
       n = 5)
```

</details>

### ANOVA

<details>

<summary>Click here for ANOVA functions</summary>

The following example data are going to be used to illustrate these functions.

```{r}
#### EXAMPLE DATA ####
set.seed(8675309)

for(i in 1:4){
  x <- rnorm(10)
  if(i == 1){
    x <- rnorm(10, mean = 2)
    data <- x |> as.data.frame()
    colnames(data) <- "Response"
    data$Explanatory <- paste0("x",i)
  }else{
    newdat <- x |> as.data.frame()
    colnames(newdat) <- "Response"
    newdat$Explanatory <- paste0("x",i)
    data <- rbind(data,newdat)
  }
}

# split into "typical" table
expanded_data <- NULL
expanded_data$x1 <- data$Response[which(data$Explanatory=="x1")]
expanded_data$x2 <- data$Response[which(data$Explanatory=="x2")]
expanded_data$x3 <- data$Response[which(data$Explanatory=="x3")]
expanded_data$x4 <- data$Response[which(data$Explanatory=="x4")]

expanded_data <- expanded_data |>
  as.data.frame()
```

The above is a *one-way ANOVA*. As a reminder, we would calculate the test as follows:

```{r}
# pivot longer does not work for one-way ANOVA, requires blocking factor
# can rbind things with same colnames to make longer

example_aov <- aov(Response ~ Explanatory, data)

summary(example_aov)
```

Above, we can see a significant result of the ANOVA. We can follow this up with a Tukey test. **This requires the package `agricolae`!** Check the ANOVA pages, however, as not all ANOVA can use this `agricolae` shortcut method.

```{r}
example_tukey <- HSD.test(example_aov,
                          # what to group by?
                          "Explanatory",
                          # significance level?
                          alpha = 0.05, 
                          # are data unbalanced
                          unbalanced = FALSE,
                          # show answer?
                          console = TRUE)
```

#### Summarize data (for plotting)

Remember - you need to change `"Explanatory"` to your explanatory variable (in quotes!) and you need to change `Response` to your response column (no quotes!). The following requires `plyr` to work, but the function itself should call up `plyr` if you do not yet have it loaded.

```{r}
# summarize by group
summary_data <- function(data, explanatory){
  require(plyr)
  ddply(data, paste(explanatory), summarise,
                 N = length(Response),
                 mean = mean(Response),
                 sd = sd(Response),
                 se = sd / sqrt(N))
}

example_summary <- summary_data(data = data, explanatory = "Explanatory")

example_summary
```

#### Significant label maker

This command requires a Tukey HSD object from `agricolae`. You can manually create a table like this for some other scenarios; see relevant pages for documentation.

```{r}
# note first group must be EXACT MATCH to your summary_data object
# groups are saved in the Tukey object
# this is true for Tukey later as well

# the following is a function that will make the significant label table
sig.label.maker <- function(tukey_test, group_name){
  sig.labels <- tukey_test$groups |> 
    # convert to a data.frame
    as.data.frame() |>
    # create a new column - place rownames into the column
    # converts to a format better for ggplot
    mutate(Explanatorys = rownames(tukey_test$groups)) |>
    # rename column to prevent confusion
    # specify dplyr; default function may be from plyr and not work
    dplyr::rename(Significance = groups)
  colnames(sig.labels)[which(colnames(sig.labels) == "Explanatorys")] <- group_name
  return(sig.labels)
}

# Function requires explanatory groups in quotes
sig_labels <- sig.label.maker(example_tukey, "Explanatory")

sig_labels
```

#### ANOVA plotter

The following function plots ANOVAS if you have a `summary_data` object and a `sig_labels` object, as shown above. **This does not work on ANOVA with interactive components**.

```{r}
anova_plotter <- function(summary_data, explanatory, 
                          response, sig_labels,
                          y_lim=NA, label_height=NA, 
                          y_lab=NA, x_lab=NA){
  require(tidyverse)
  
  plot_data_1 <- summary_data[,c(explanatory, response, "se")]
  plot_data_2 <- sig_labels[,c(explanatory,"Significance")]
  
  colnames(plot_data_1) <- c("explanatory", "response", "se")
  colnames(plot_data_2) <- c("explanatory", "Significance")
  
  plot_data <- plot_data_1 |> 
    full_join(plot_data_2, by = "explanatory")
  
  if(is.na(y_lim)){
    if(min(plot_data$response) < 0){
      y_lim <- c(min(plot_data$response) - 
          4*max(plot_data$se),
        max(plot_data$response) + 
          4*max(plot_data$se))
    }else{
      y_lim <- c(0,max(plot_data$response) + 
                                4*max(plot_data$se))
    }
  }
  if(is.na(label_height)){label_height <- 0.25*max(y_lim)}
  if(is.na(y_lab)){y_lab <- "Response"}
  if(is.na(x_lab)){x_lab <- "Treatment"}
  
  plot_1 <- ggplot(plot_data,
         aes(x = explanatory, y = response)) +
    geom_point() +
    geom_errorbar(data = plot_data,
                  aes(ymin = response - 2*se,
                      ymax = response + 2*se,
                      width = 0.1)) +
    ylim(y_lim) +
    theme_classic() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, size = 5)) +
  geom_text(data = plot_data,
            # make bold
            fontface = "bold",
            # define where labels should go
            aes(x = explanatory, 
                # define height of label
                y = label_height, 
                # what are the labels?
                label = paste0(Significance))) +
  xlab(x_lab) +
  ylab(y_lab)
  
  print(plot_1)
}

anova_plotter(summary_data = example_summary,
              explanatory = "Explanatory",
              response = "mean", # from summary_data table! What is to be plotted
              sig_labels = sig_labels)
```

Note that in the above, the default label height is not working for us. We can adjust this with `label_height`.

```{r}
anova_plotter(summary_data = example_summary,
              explanatory = "Explanatory",
              response = "mean", # from summary_data table! What is to be plotted
              sig_labels = sig_labels,
              label_height = -1)
```

Much better!

</details>
