---
title: "Glossary"
author: "Dr. Jacob C. Cooper"
format: html
editor: visual
---

## Common Commands

```{r, echo = F, eval = T, message = F}
library(tidyverse)
```

The following are common useful commands used in *R*, with examples of their use.

-   `<-` / `=` - save a value as an object

```{r}
x <- 10
x
```

-   `%>%` - "pipe" a command or output into another command. Required `tidyverse` to run, and you can use the shortcut `CTRL SHIFT M` on Windows or Mac.

```{r}
# make repeatable
set.seed(930)

# random string
x <- rnorm(20)

x %>% 
  # pass to summary
  summary() %>% 
  # pass summary through round
  round(2)
```

-   `c` - concatenate, place two values together

```{r}
x <- c(10,11)
x
```

## Basic statistics

For these examples, we will create a random vector of number to demonstrate how they work.

```{r}
x <- rnorm(1000)
```

-   `mean` - get the mean / average of a set of data

```{r}
mean(x)
```

## Picking the test

For a step by step click-through, see below:

Is the explanatory variable categorical or continuous?

### Categorical explanatory variables

Is the response variable categorical or continuous?

#### Categorical explanatory & response variables

When both variables are categorical, we are looking at $\chi^2$ tests where we are comparing counts of individuals from the two different categories. Our major types of tests are:

-   $\chi^2$ goodness-of-fit tests - looking at data or counts for different categories and asking if what we *observe* is the same as what we *expect*

-   $\chi^2$ tests of independence -

-   Fisher's exact test

#### Categorical explanatory and continuous response variables

### Continuous explanatory variables

#### Continuous explanatory variables and categorical response variables

#### Continuous explanatory & response variables
