---
title: "t_test_1_samp_HW"
output: html_document
---

### Answer each question. Perform all necessary tests. Perform transformations on the data if required.

```{r}
# install.packages("lme4")   # required for "sleepstudy" dataset
library(lme4)
data("sleepstudy")
head(sleepstudy)

# install.packages("MASS")   # required for "galaxies" dataset
library(MASS)
data("galaxies")
head(galaxies)

```

### **Question 1:**

These are the test scores of 12 students on a quiz. The quiz was out of 100 points. Suppose we want to test whether the mean score differs significantly from a predicted average of 70.

```{r}
quiz_scores <- c(75, 82, 68, 90, 73, 85, 77, 79, 88, 91, 83, 80)  
```

-   State the **null and alternative hypotheses**.

$H_0:\mu=70$
$H_A:\mu\ne70$

-   Are these data normal?

```{r}
shapiro.test(quiz_scores)
```

Yes these are normal.

```{r}
t.test(x = quiz_scores, mu = 70)
```


-   Is the mean score significantly different from the expected result?

Yes, $p = 0.0002$.

-   Did the students do better or worse than expected, if there is a difference?

Better, value is higher.

### **Question 2:**

The following is a list of reported study hours for Biostats per week. We expect the class average to be about three hours a week. Using this dataset, answer the following questions:

```{r}
study_hours <- c(0.5, 3.0, 2.5, 4.5, 3.0, 1.5, 2.0, 3.5, 6, 1.0)  
```

-   State the **null and alternative hypotheses**.

$H_0:\mu=3$
$H_A:\mu\ne3$

-   Are these data normal?

```{r}
shapiro.test(study_hours)
```

```{r}
t.test(study_hours, mu = 3)
```


-   Do students spend the expected amount of time studying per week?

Yes, the spend 3 hours on average studying.

-   Do students spend more or less time studying per week, if there is a difference?

There is no significant difference; mean is slightly lower.

### **Question 3:**

The following dataset records the reaction times of people who have had less than three hours of sleep on the night before this test. Using the reaction time column, perform a $t$-test to determine if these people have a statistically different reaction time than the human average (250 ms).

```{r}
reaction_times <- sleepstudy$Reaction
```

-   State the **null and alternative hypotheses**.

$H_0:\mu=250$
$H_A:\mu\ne250$

-   Are these data normal?

```{r}
shapiro.test(reaction_times)
```

These data are not normal.

```{r}
shapiro.test(log(reaction_times+1))
```

Log transformation needed.

```{r}
t.test(log(reaction_times + 1), mu = log(250 + 1))
```

-   Is the mean score significantly different from the expected result?

Yes, the score is different. $p < 0.0001$.

-   Are the people in the dataset slower or faster than average, if there is a difference? What might be the reason for this?

The reaction time is slower; `r exp(5.685031)-1`.

### **Question 4:**

Whole milk is expected to be around 3.25% fat. Researchers from Florida wanted to determine if this was the case and used two methods to measure the fat percentage in the milk they tested. Using the enzymatic method (\$triglyceride), determine if the fat percentage of this milk was significantly different from the 3.25% expected.

```{r}
milk <- read.csv("https://users.stat.ufl.edu/~winner/data/milkfat_measure.csv")
milk_fats <- milk$triglyceride

milk_fats
```

-   State the **null and alternative hypotheses**.

$H_0:\mu=3.25$
$H_A:\mu\ne3.25$

-   Is the mean score significantly different from the expected result?

```{r}
shapiro.test(milk_fats)
```

```{r}
t.test(milk_fats, mu = 3.25)
```

-   Is the milk fattier or leaner than expected, if there is a difference?

Leaner; p = 0.02, mu = 2.80.

### **Question 5:**

Galaxies are rapidly moving away from us at various speeds. Previous studies had offered an average recession rate of 20,000 km/s. Data collected using redshift allows us to calculate the actual speed of recession of a galaxy. Using the data from R. J. Roeder (1990), saved as "galaxies", determine if the average galaxy is actually receding at the previously estimated rate.

```{r}
head(galaxies)
```

-   State the **null and alternative hypotheses**.

$H_0:\mu=20,000$
$H_A:\mu\ne20,000$

-   Is the mean score significantly different from the expected result?

```{r}
shapiro.test(galaxies)
```

```{r}
shapiro.test(sqrt(galaxies))
```

Not normal! No transformation works.

```{r}
wilcox.test(galaxies, 20000)
```

-   Are the galaxies moving away faster or slower, if there is a difference?

No, they are not. p = 0.69.