---
title: "9_prob_distributions"
author: "UNK Bio"
date: "`r Sys.Date()`"
output: html_document
---

# 8.6.1 Problem 1

There is a claw machine at the local fair that is notoriously difficult to use. In fact, it is estimated that only 1 in 50 attempts are successful at getting a toy out of the machine. What is the probability that you would try it 50 times and win thrice?

```{r}
binom.test(x = 3, n = 50, p = 1/50)
```

The probability is low - 7.9% (round to 8%).

# 8.6.2 Problem 2

Whooping Cranes *Grus americanus* are some of the most endangered birds in North America, with only 557 birds in the population that migrates through Nebraska. Conversely, approximately 736,000 Sandhill Cranes *Antigone canadensis* migrate through Nebraska. Assuming that census counts of flocks of 5000 cranes have the expected average ratio of cranes, what are the odds of finding a crane flock with two or more Whooping Cranes?

```{r}
lower <- sum(dbinom(x = 0:1, size = 5000, prob = (557/736000)))

(1-lower) |> round(2)
```

Probability is 89%.

# 8.6.3 Problem 3

Your friend has a twenty-sided dice that you suspect to be unfair. They have rolled a 20 four times in the past thirty rolls. Is the dice unfair?

```{r}
binom.test(x = 4, n = 30, p = 1/20)
```

No, the dice may be fair.

# 8.6.4 Problem 4

You roll three die at a single time. What are the odds of getting exactly two fours?

```{r}
dbinom(x = 2, # two rolls
       size = 3, # three die at once
       1/6) |> # 1/6 chance on each die 
  round(2) # round answer
```

Approx. a 7% chance.

# 8.6.5 Problem 5

You notice that in a class of 200 people, the odds of any given student being absent is approximately 3%. What are the odds that 5 people miss a given class?

```{r}
dbinom(x = 5,
       size = 200,
       prob = 3/100) |> 
  round(2)
```

Approximately 16%.

# 8.6.6 Problem 6

There are approximately 3 severe storm warnings per 16 week semester at UNK. What are the odds of having a semester with no severe storm warnings?

```{r}
dbinom(x = 0, size = 16, prob = 3/16)
```

Around 3.6%.

# 8.6.7 Problem 7

You are giving mice an experimental treatment that is designed to change their fur color. In a trial of 500 mice, 15 had their fur change color. Using this precedent, are the odds of having 6 mice have their fur change color in another trial of 500 individuals?

```{r}
dbinom(x = 6, size = 500, prob = 15/500)
```

# 8.6.8 Problem 8

What are the odds of rolling no fours on two twenty-sided die?

```{r}
# odds of not getting a four
success <- dbinom(x = 0, size = 2, prob = 1/20)

print(success)
```

90.25% chance of not rolling that.

# 8.6.9 Problem 9

Dr. Cooper once caught his mom shaking sodas and putting them back at the shelves at a gas station while they were traveling. Assuming his mom was able to shake 10% of the sodas before getting caught, what are the odds of getting three or more shaken sodas if you buy 20 sodas?

```{r}
1 - sum(dbinom(x = 0:2, size = 20, prob = 0.1))
```

32% chance.
