---
title: "Chapter 4 Key"
author: "UNK Biology"
date: "`r Sys.Date()`"
output: html_document
---

### Homework instructions

Please create an *RMarkdown* document that will render as an `.html` file. You will submit this file to show your coding and your work. Please refer to the [Introduction to *R*](https://jacobccooper.github.io/biol105_unk/intro_to_r.html) for refreshers on how to create an `.html` document in *RMarkdown*. You will need to do the following for each of these datasets:

-   mean

-   median

-   range

-   interquartile range

-   variance

-   standard deviation

-   coefficient of variation

-   standard error

-   whether there are any "outliers"

**Please show all of your work for full credit.**

### Data for homework problems

Please use the following datasets for your homework.

```{r}
library(tidyverse)

se <- function(x){
  n <- length(x) # calculate n
  s <- sd(x) # calculate standard deviation
  se_val <- s/sqrt(n)
  return(se_val)
}

cv <- function(x){
  sigma <- sd(x)
  mu <- mean(x)
  val <- sigma/mu
  return(val)
}
```

*Note* when R rounds it will cut off trailing zeroes... now how many decimals points you need!

#### Problem 4.1

```{r}
x <- c(2,5,3,7,8,3,9,3,10,4,7,4,6,11,9,
       9,11,5,7,3,8,9,2,1,3,8,3,8,9,3)

key_maker <- function(x,roundval){
  print(paste0("Mean: ", round(mean(x),roundval)))
  
  print(paste0("Median: ", round(median(x),roundval)))
  
  print(paste0("Range: ", round(max(x) - min(x),roundval)))
  
  quantiles_x <- quantile(x, type = 6) %>% as.numeric()
  print(paste0("IQR: ",round(quantiles_x[4] - quantiles_x[2],roundval)))
  
  print(paste0("Variance: ",round(var(x),roundval)))
  
  print(paste0("SD: ", round(sd(x),roundval)))
  
  print(paste0("CV: ",round(cv(x),roundval)))
  print(paste0("SE: ", round(se(x),roundval)))
  
  lowquant <- quantile(x,0.25,type = 6) %>% as.numeric()
  hiquant <- quantile(x,0.75,type = 6) %>% as.numeric()
  
  iqr <- hiquant - lowquant
  
  lowbound <- mean(x) - (1.5*iqr)
  hibound <- mean(x) + (1.5*iqr)
  
  # low outliers?
  # select elements that match
  # identify using logical "which"
  lows <- sum(x < lowbound)
  if(lows > 0){
    low.vals <- x[which(x < lowbound)]
    print("Low outlier(s): ")
    print(x[which(x < lowbound)])
  }else{print("No low outliers.")}
  highs <- sum(x > hibound)
  if(highs > 0){
    hi.vals <- x[which(x > hibound)]
    print("High outlier(s):")
    print(hi.vals)
  }else{print("No high outliers.")}
}
```

```{r}
key_maker(x,roundval = 0)
```

#### Problem 4.2

Dataset on culmen (top of bill) lengths in Belted Kingfishers *Megaceryle alcyon*.

```{r}
# beki = Belted Kingfisher (Megaceryle alcyon)
beki <- c(48.1, 50.8, 48.8, 56.8, 57.7, 47.0,
          56.8, 60.2, 55.8, 59.2, 52.5, 50.4,
          48.0, 57.1, 51.8, 52.3, 47.8, 58.0,
          53.4, 55.2, 51.0, 59.3, 61.5, 61.2,
          57.8, 50.1, 56.0, 56.5, 55.8, 56.5,
          56.3, 59.8, 61.8, 56.2, 57.5, 59.3,
          62.4, 61.1, 59.9, 55.6, 56.8, 59.2)

key_maker(beki,roundval = 1)
```

#### Problem 4.3

Dataset on Great Tit brood sizes.

```{r}
# brood size in Great Tits (Parus major)
# must run this whole box to get it formatted correctly
parus_major <- matrix(nrow = 16, ncol = 2, byrow = T,
                      data = c(1, 5,
                               2, 10,
                               3, 32,
                               4, 61,
                               5, 81,
                               6, 95,
                               7, 80,
                               8, 88,
                               9, 72,
                               10, 67,
                               11, 61,
                               12, 41,
                               13, 24,
                               14, 1,
                               15, 1,
                               16, 2)) %>%
  as.data.frame()

colnames(parus_major) <- c("Brood_size", "Frequency")

# need to expand the table
# thanks Stack Exchange!

parus_expanded <- parus_major[rep(1:nrow(parus_major), parus_major[["Frequency"]]), ]

key_maker(parus_expanded$Brood_size)
```