---
title: "t_test_1_samp_HW"
output: html_document
---

### Answer each question. Perform all necessary tests. Perform transformations on the data if required.

```{r}
install.packages("lme4")   # required for "sleepstudy" dataset
library(lme4)
data("sleepstudy")
head(sleepstudy)

install.packages("MASS")   # required for "galaxies" dataset
library(MASS)
data("galaxies")
head(galaxies)

```

### **Question 1:**

These are the test scores of 12 students on a quiz. The quiz was out of 100 points. Suppose we want to test whether the mean score differs significantly from a predicted average of 70.

```{r}
quiz_scores <- c(75, 82, 68, 90, 73, 85, 77, 79, 88, 91, 83, 80)  
```

-   State the **null and alternative hypotheses**.

-   Is this data normal?

-   Is the mean score significantly different from the expected result?

-   Did the students do better or worse than expected, if there is a difference?

### **Question 2:**

The following is a list of reported study hours for Biostats per week. We expect the class average to be about three hours a week. Using this dataset, answer the following questions:

```{r}
study_hours <- c(0.5, 3.0, 2.5, 4.5, 3.0, 1.5, 2.0, 3.5, 6, 1.0)  

```

-   State the **null and alternative hypotheses**.

-   Is this data normal?

-   Do students spend the expected amount of time studying per week?

-   Do students spend more or less time studying per week, if there is a difference?

### **Question 3:**

The following dataset records the reaction times of people who have had less than three hours of sleep on the night before this test. Using the reaction time column, perform a T-test to determine if these people have a statistically different reaction time than the human average (250 ms).

```{r}
reaction_times <- sleepstudy$Reaction

```

-   State the **null and alternative hypotheses**.

-   Is this data normal?

-   Is the mean score significantly different from the expected result?

-   Are the people in the dataset slower or faster than average, if there is a difference? What might be the reason for this?

### **Question 4:**

Whole milk is expected to be around 3.25% fat. Researchers from Florida wanted to determine if this was the case and used two methods to measure the fat percentage in the milk they tested. Using the enzymatic method (\$triglyceride), determine if the fat percentage of this milk was significantly different from the 3.25% expected.

```{r}
milk <- read.csv("https://users.stat.ufl.edu/~winner/data/milkfat_measure.csv")
milk_fats <- milk$triglyceride

milk_fats

```

-   State the **null and alternative hypotheses**.

-   Is this data normal?

-   Is the mean score significantly different from the expected result?

-   Is the milk fattier or leaner than expected, if there is a difference?

### **Question 5:**

Galaxies are rapidly moving away from us at various speeds. Previous studies had offered an average recession rate of 20,000 km/s. Data collected using redshift allows us to calculate the actual speed of recession of a galaxy. Using the data from R. J. Roeder (1990), saved as "galaxies", determine if the average galaxy is actually receding at the previously estimated rate.

```{r}
head(galaxies)
```

-   State the **null and alternative hypotheses**.

-   Is this data normal?

-   Is the mean score significantly different from the expected result?

-   Are the galaxies moving away faster or slower, if there is a difference?
